### 2022-11-10 submission/manuscript.Rmd
Change coverage from fraction to percent

* Rarefy your data
* Problems with recommended methods...
  * Many recommended methods are borrowed from gened expression analysis
  * Meaning of zeroes in data - structural vs. below limit of detection
* Factors that determine what number of sequences to rarefy to
* Need better methods of pooling libraries that result in more even distribution of sequences across samples
* Rarefy your data

>>>>

Over the past decade, the question of whether to rarefy microbial community sequence data has been controversial. The analyses I presented here strongly indicate that rarefaction is necessary to control for uneven sampling effort when comparing communities based on alpha and beta diversity indices. Compared to all other methods, rarefaction removed the correlation between sequencing depth and alpha or beta diversity metrics when the sequencing depth varied by as much as 97-fold across samples. I showed that this correlation could lead falsely detecting differences between treatment groups if sampling depth and sequencing effort are confounded [e.g., @Morris2013]. The correlation with sampling effort leads to an artificial increase in the variation between samples and a reduced power to detect true differences in alpha and beta diversity. For these reasons, rarefaction is a valuable tool to control for uneven sampling effort until improved statistical procedures are developed or it becomes possible to more evenly distribute sequencing effort across samples.

The primary alternative to rarefaction that many advise is to estimate alpha diversity has been to use non-parametric or parametric methods using raw data and to then compare the estimates [@Willis2019]. My data demonstrates that such approaches are limited for several reasons. First, it has been widely observed that non-parametric richness esimates such as ACE and Chao1 are sensitive to sampling effort. Therefore, these strategies do not, in practice, remove the effects of sampling effort. Second, parametric approaches, such as those implemented in the breakaway R package, generate confidence intervals that are likely to include the true richness and theoretically shrink with increased sampling effort. For most samples, the confidence intervals around the estimates are too wide to be meaningful, again leading to an inability to remvoe the effects of sampling effort. Third, it has become an increasingly common practice for researchers to remove sequences that are rare in a sample (e.g., those that appear once). Although that approach was not taken in this study, removing rare sequences would skew the distribtion of sequences and OTUs leading to a distortion of the measurement of alpha diversity [@Schloss2020]. The effects of removing rare sequences would vary across samples depending on the number of sequences in each sample. One interesting result of this analysis was the demonstration that as metrics that depend less on rare taxa are used, the effect of uneven sampling effort was reduced. For example, richness counts a sequence appearing once as much as sequence appearing 1,000 times, while Shannon diversityindex places more emphasis on the more abundant sequence, and the inverse Simpson index even more. Although normalizing communities to a common number of sequences is also suggested (e.g., SRS normalization) to control for uneven sampling effort, the current analysis indicates that its performance does not meet that of rarefaction. For analysis of alpha diversity metrics, rarefaction is the most effective and consistent approach to control fo uneven sampling effort.

Use of relative abundances, normalized counts, variance stabilizing transformations, and centered log ratios have each been recommended as more robust alternatives to rarefaction. Again, the only approach that consistently removed the effects of uneven sampling effort was rarefaction. Most of the recommendations borrow techniques from methods for identifying differential gene expression. Unfortunately, there is an important but underappreciated difference between gene expression and community data. This is the interpretation of unobserved data. For gene expression analysis in a single organism the lack of any sequencing data for a gene would indicate that its expression was below the limit of detection. Sequencing the same organism under multiple conditions would not lead to a seemingly unbouned number of genes in the organims. Rather, the number of genes has a definite limit that is knowable from the genome sequence. With microbiome data, an unobserved sequence could mean that the organism was present, but below the limit of detection or that the organism was missing. Because we have yet to exhaustively sample any community in the same way we have sequenced a single genome it is unreliable to impute the presence of all organisms. Yet, this is exactly what the variance stabilization transformation and most CLR techniques do. This analysis has demonstrated a clear correlation between distances calculated by these methods and sequencing effort. This result is at odds with the claims by others that the distances are scale invariant [@Martino2019]. Again, rarefaction is the most effective and consistent approach to control for uneven sampling effort when calculating beta diversity metrics.

Two common critique of rarefaction is that the approach ommits valid data and that the selection of the value to rarefy the sequence data to is arbitrary [@McMurdie2014]. I disagree with both critiques. All of the data are used to calculate the mean value of the rarefied metrics; none of it is excluded. When the dataset is subsampled, every sequence has a random chance of being included in the calculated metric. When that subsampling is repeated a large number of times (e.g., 100 or 1000) the risk of oversampling rare taxa is mititgated. As for the second critism, I would resist the claim that the selection of the rarefaction threshold is arbitrary. In practice, there is a tradeoff between sampling breadth and depth. Greater breadth will increase the statistical power to compare treatment groups and greater depth will increase the resolution to describe the communities. My process for picking a break involves looking for a natural break in the distribution of the number of sequences. For example, the Lake dataset used in this study had a clear break around 10,000 sequences. I would also consider what samples are below any break that I select. If there were critical samples below the break I would either reduce the break point or I would generate more sequences from those samples. Furthermore, regardless of the break point one selects, they could repeat an analysis at multiple thresholds to assess the tradeoff between statistical power and resolution. As shown in my analysis of Good's coverage values, most studies obtain an ample level of coverage and would need to increase their sampling depth by 10 fold to increase the coverage by several percentage points. I contend that for most studies increased sequencing effort should be applied to improve sampling breadth and statistica power over depth and resolution.

* Need better methods of pooling libraries that result in more even distribution of sequences across samples
* We don't have true absolute numbers
* Rarefy your data
